Selective Sequential Model Selection

### 2015-10-13 paper.tex
major reorg of paper

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\tL}{\widetilde{L}}
\newcommand{\hk}{\hat{k}}
\newcommand{\hr}{\hat{r}}       
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}
\newcommand{\Err}{\mathcal{E}}

@@

\title{Adaptive Sequential Model Selection}
\author{William Fithian, Jonathan Taylor, Rob Tibshirani, and Ryan Tibshirani}
\maketitle

\begin{abstract}
  Many model selection algorithms produce a ``path'' of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path, accounting for the fact that the model path is determined adaptively using the data. In the case of linear regression, our $p$-values improve on the power of the spacings test of \citet{taylor2014exact}, often dramatically.
  We also show that the selected models test  bears a close relationship to parametric bootstrap sampling.

To choose a model we feed the resulting $p$-values as inputs into sequential stopping rules proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving adaptive control of the familywise error rate or false discovery rate. These stopping rules assume that the single-step null $p$-values are independent of each other and of the non-null $p$-values, a condition that bmiis not satisfied by the $p$-values of \citet{taylor2014exact}. We derive intuitive and general conditions for independence and show that our proposed constructions yield independent $p$-values.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Many model selection procedures produce  a sequence of increasingly complex models. Often our goal is to choose the simplest model that is not falsified by the available data.
In other words, we seek  the index of the smallest adequate model ---  the smallest model that cannot be falsified given available evidence.
Examples of such model selection procedures include  forward stepwise linear regression, least angle regression (LAR) and  the ever-active path in $\ell_!$-regularized models .
All of these were studied in \citet{taylor2014exact}. Another, different, example is principal components analysis in which we try to determine the correct rank of the
covariance matrix. This problem was studied in \citet{choi2014selecting}.

A natural way carry out this model selection is as follows: we compute a set  of sequential p-values   at each step of the path of models and then feed them into a stopping rule
that provides guaranteed control of the false discovery rate (FDR) or a similar quantity. Recently  \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind.
Both sets of rules require that once we have reached the correct model in our path, the p-values in subsequent steps are uniform (or sub-uniform)
and independent.
In this paper we develop a theoretical framework for the construction of sequential p-values  that are both (sub)-uniform and independent under the correct model,
and show how to construct them  in forward stepwise regression, LAR and the ever-active path in $\ell_1$-regularized models.
In the former case, the test is simply a sequential selective version the max-$t$ test of ~\citet{buja2014}, based on the theory developed \citet{fithian2014optimal}.

Table \ref{tab:diab} shows an example: using the diabetes data taken from \cite{LARS}.
These data consist of  observations on $442$ patients, with the response of interest being a quantitative measure of disease progression one year after baseline.
There are ten baseline variables---age, sex, body-mass index, average blood pressure,
and six blood serum measurements---plus quadratic terms, giving a total of 64 features.

\begin{table}[ht]
\centering
\input{tables/diabetes_edit.tex}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive p-values from forward stepwise regression, 
p-values from the ``saturated'' model, and the ``max-T'' p-values (computed without knowing $\sigma^2$) that are presented in this paper. The red annotation indicates the stopping step
prescribed by {\tt forwardStop}. For the max-T p-values, this stopping rule gives exact $\text{FDR}_{\text{model}}$ control at the 10\% level.}
\label{tab:diab}
\end{table}

The table shows the naive p-values from forward stepwise regression, that do not account for the selection, as well as the
p-values from the ``saturated'' model, and the ``max-t'' p-values that are presented in this paper.
The latter two sets of p-values test someone different hypotheses:  the p-values from the saturated model test whether the partial regression of the predictor just added is zero,
what we call the ``incremental null''
In contrast, the max-t pvalues test whether the current model is adequate, that is, all predictors not in the model  have coefficients equal to zero,
the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
Now it turns out that the max-t test is often more powerful that the saturated model test, and importantly, the max-t pvalues are exactly
independent after we reach the point the model sequence when the complete null is true. This independence means that we can apply
sequential FDR exact stopping rules, such as ``forwardStop'' (\citet{gsell2013sequential}). When applied to the max-t pvalues at FDR 0.10, {\tt forwardStop}
suggested stopping at step 8. In constant, the saturated lvalues gave stopping a point of 3, with no guarantee of FDR control
due to the lack of independence of the p-values.

An outline of this paper is as follows:
In Section \ref{sec:genericSetting} we setup the problem and give some important definitions. In Section \ref{sec:sequential} we review recently proposed methods for
testing ordered hypothesis using from a sequence of p-values. These method require that the p-values be uniform and independent under the null, and we derive general conditions
to ensure this property. In Section \ref{sec:singleStep}  we  show how to construct valid selective $p$-values for a single step in the path and derive the max-t test
for  forward stepwise regression and the Next-$\lambda$ test for the lasso. Section \ref{sec:computation}  describes our computational strategy for estimating the p-values using accept-reject and hit and run sampling. We also relate the method to parametric bootstrap sampling. We provide a simulation study in Section \ref{sec:sparseReg} and briefly discuss applications of our framework to
principal components analysis in section\ref{sec:pca}. The paper ends with a Discussion.

%\begin{example}[Forward-Stepwise Linear Regression]
 % \citet{taylor2014exact}
%end{example}

%\begin{example}[The LARS Algorithm in Regression]
%  \citet{taylor2014exact}
%\end{example}

%\begin{example}[Ever-Active Path in $\ell_1$-Regularized Methods]
 % \citet{taylor2014exact}
%\end{example}

%\begin{example}[Principal Components Analysis]
%  As a second motivating example, consider model selection for   principal components analysis. In that case we are given a data matrix $X \in \R^{n\times d}$, with which we form a sample %covariance matrix
%\[
%S = \frac{1}{n-1} \sum_{i=1}^n(x_i - \bar x)^2
%\]
%The first $d$ principal component loadings are the first $d$ eigenvectors of $S$, which call $u_1,\ldots, u_d$. These induce a sequence of nested Wishart models:
%\[
%M_0 \sub M_1 \sub \cdots \sub M_d
%\]
%n which
%\begin{equation}
 % M_k:\; (n-1) S \sim W_d\left(\lambda_0 I_d + \sum_{i=1}^k     \lambda_i u_i u_i', \;\;\; n-1\right).
%\end{equation}
%This problem was studied in \citet{choi2014selecting}.
%\end{example}

\section{Notation and Problem Setting}
\label{sec:genericSetting}

More generically, we observe data $Y \in \cY$, with unknown sampling distribution $F$. We then use $Y$ to generate an adaptive sequence of $d$ nested models
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
\WFcomment{should probably end with $M_{d-1}(Y)$. need to fix this throughout paper.}
\RTcomment{Will- be more specific here---   A parametric model  with
  specified nonzero support?}
\WFcomment{Not necessarily. As far as I can tell, there's no need to restrict ourselves in either case. $Y\sim U[0,\theta]$ is a perfectly good model; so is a Cox proportional hazards model.}

By ``model,'' we mean a family of probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:k}$ to denote the sequence of the first $k$ models.

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. Our goal is to examine the data and output a {\em stopping rule}, i.e. an estimator $\hk(Y)$ of $k_0$. We consider $M_k$ to be ``rejected'' if $k < \hat k$, and ``accepted'' otherwise. Thus $\hk$ is the number of models we rejected, while $k_0$ is the number that we should have rejected.

The number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Depending on the context we might want to control the familywise error rate (FWER) $\P(V>0)$, the false discovery rate (FDR) $\E[V/(\hk \vee 1)]$, or another error rate given by the expectation of some loss function $g$:
\begin{equation}\label{eq:errRate}
\Err_F(\hk(\cdot), g) = \E_F\left[ g\left(\hk(Y), k_0(Y)\right)\right]
\end{equation}

We will assume that for each candidate model $M$ there is a minimal sufficient statistic $T(Y; \,M)$, and write
\[
T_k(Y) = T(Y; M_k(Y)).
\]

First suppose that we have single step p-values for testing
\[
H_{k}:\; F\in M_{k-1}(Y)
\]
against the alternative that $F\in M_\infty\setminus M_{k-1}$. The $p$-values must account for the fact that the models are chosen adaptively. 
\citet{gsell2013sequential} and \citet{li2015accumulation} propose stopping rules that operate on the sequence $p_{1:d}$. These stopping rules provably control certain error rates, provided that the null $p$-values are mutually independent and independent of the non-null $p$-values. 
We review these methods in Section~\ref{sec:sequential} .

Hence the task is to  construct sequential p-values which satisfy the above requirements, and we do this in Section \ref{sec:singleStep}.
We also discuss general conditions under which single-step $p$-values are independent.

\subsection{Linear Regression}

Most of our examples will focus on the important and familiar problem of adaptively selecting a set of predictors in linear regression models. In this problem, we observe a random response $Y\in \R^n$ as well as a fixed design matrix $X\in \R^{n \times p}$, whose columns correspond to candidate predictors. For each {\em active set} $E \sub \{1,\ldots,p\}$, there is a corresponding candidate model
\[
M(E):\; Y \sim \cN( X_E\beta, \sigma^2 I_n),
\]
which is a subset of the {\em full model}
\[
M_\infty:\; Y \sim \cN(X\beta, \sigma^2I_n).
\]
The error variance $\sigma^2$ may be known or unknown depending on the context. The complete sufficient statistic for $M(E)$ is $X_E'Y$ if $\sigma^2$ is known, or else $\left(X_E'Y, \|Y\|^2\right)$.

@@

All of the specific examples considered in this paper will have a common form generalizing the linear regression problem above. Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^{\cJ}$:
\[
M_\infty = \{F_\theta:\; \theta \in \Theta\}.
\]
For any subset $E\sub \cJ$ define the sparse submodel with active coefficients $E$ as follows:
\[
\Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
\quad M(E) = \{F_\theta:\; \theta\in \Theta(E)\}.
\]

This setting includes linear regression problems with known or unknown $\sigma^2$ as special cases. A typical path algorithm returns a sequence of nested active sets 
\[
E_0(Y) \sub E_1(Y) \sub \cdots \sub E_d(Y),
\]
inducing the model path given by $M_k = M(E_k)$.

As an example, in forward stepwise regression with known $\sigma^2$, we might include an unknown intercept term $\beta_0$ in the global null model $M_0$, then at each step $k\geq 1$, add the predictor $j_k$ that reduces the residual sum of squares the most. In that case $E_0 = \{0\}$, and $E_k = E_{k-1} \cup \{j_k\}$ for $k\geq 1$.

\subsection{Path Algorithms}

We will be especially interested in two path algorithms in the sparse parametric setting: forward-stepwise paths and ever-active regularization paths.

\paragraph{Forward Stepwise Paths}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. The {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed $E_0$, then at step $k=1,\ldots,d$, we define
\begin{align}\label{eq:forwardDef_start}
j_k &= \argmax_j \;\;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\} \\
E_k &= E_{k-1} \cup \{j_k\}\\\label{eq:forwardDef_end}
M_k &= M(E_k).
\end{align}
That is, at each step we add to the model the one variable that would most increase the likelihood. In the case of linear regression, this amounts to adding the variable that maximally reduces the residual sum of squares.

\paragraph{Ever-Active Regularization Paths}
Another class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. For $r=0,1,\ldots,m$, let $P_r(\theta)$ denote some regularization penalty, and define
\begin{align}\label{eq:regPathDef_start}
  \hat\theta^{r}(Y) &= 
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_r(\theta) \\
  \tE_r(Y) &= \left\{j:\; \hat\theta_j^s \neq 0 
    \text{ for any } s \leq r \right\}
\end{align}

While the sets $\tE_r$ are nested by definition, we could have $\tE_r = \tE_{r+1}$ for most values of $r$. Thus, we will take the sequence of distinct ever-active sets. Formally, let $R_0=0$, and for $k\geq 1$ let $R_k$ denote the (random) index where the active set actually changes for the $k$th time. That is,
\begin{align}
  R_k &= \min\{s:\; \tE_s \neq \tE_{R_{k-1}}\}\\
  E_k &= \tE_{R_k}\\
  \label{eq:regPathDef_end}
  M_k &= M(E_k)
\end{align}

The ever-active lasso path in linear regression is the most familiar example, in which we take $P_r(\beta) = r^{-1}\|\beta\|_1$, and the penalized log-likelihood criterion is
\begin{equation}
  \argmin \frac{n}{2}\log(2\pi\sigma^2) 
  + \frac{1}{2\sigma^2}\|Y - X\beta\|_2^2 + \frac{1}{r}\|\beta\|_1
\end{equation}
If $\sigma^2$ is known, then $\hat\beta^{r}$ is the lasso solution with Lagrange parameter $\lambda^r = \sigma^{2}/r$. If $\sigma^2$ is unknown, it is the lasso solution for $\lambda^r = (\hat\sigma^r)^2/r$. Because $\hat\sigma^r = n^{-1/2}\|Y-X\hat\beta^r\|_2$ is decreasing in $r$, $\lambda^r$ is strictly decreasing in $r$.

Because it is actually possible to solve the lasso at every value of the Lagrange parameter, it is not really necessary that $r$ only take on integer values. We can instead notionally take $r\in (0,\infty)$, and use a path algorithm like LARS \citep{taylor2014exact} to find every point in the path where the lasso solution changes.

\subsection{``Model-centric'' and ``variable-centric'' null hypotheses}
\label{sec:whichnull}

\WFcomment{Shorten this section, move some of it to Discussion?}

>>>>

\newcommand{\cM}{\mathcal{M}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\FDR}{\textnormal{FDR}}
\newcommand{\FCR}{\textnormal{FCR}}
\newcommand{\crt}{\phi}
\newcommand{\M}{\mathcal{M}}
\newcommand{\cY}{\mathcal{Y}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\cV}{\mathcal{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\Gv}{\;\;\big|\;\;}
%\newcommand{\cP}{\mathcal{P}}
\newcommand{\proj}{\cP}
\newcommand{\pow}{\text{Pow}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\cF}{\mathcal{F}}
\newcommand{\sC}{\mathscr{C}}
\newcommand{\hJ}{\widehat{J}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\tM}{\widetilde{M}}
\newcommand{\tE}{\widetilde{E}}
\newcommand{\tV}{\widetilde{V}}
\newcommand{\tR}{\widetilde{R}}
\newcommand{\tL}{\widetilde{L}}
\newcommand{\hk}{\hat{k}}
\newcommand{\hr}{\hat{r}}       
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cJ}{\mathcal{J}}
\newcommand{\leqAS}{\overset{\textrm{a.s.}}{\leq}}
\newcommand{\Err}{\mathcal{E}}
\newcommand{\RSS}{\text{RSS}}

@@

\title{Selective Sequential Model Selection}
\author{William Fithian, Jonathan Taylor, Rob Tibshirani, and Ryan Tibshirani}
\maketitle

\begin{abstract}
  Many model selection algorithms produce a path of fits specifying a sequence of increasingly complex models. Given such a sequence and the data used to produce them, we consider the problem of choosing the least complex model that is not falsified by the data. Extending the selected-model tests of \citet{fithian2014optimal}, we construct $p$-values for each step in the path, accounting for the fact that the model path is determined adaptively using the data. In the case of linear regression, our $p$-values improve on the power of the spacings test of \citet{taylor2014exact}, sometimes dramatically.

To select a model, we can feed the resulting $p$-values as inputs into sequential stopping rules such as those proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, achieving control of the familywise error rate or false discovery rate as desired. Some stopping rules require the null $p$-values to be independent of each other and of the non-null $p$-values, a condition not satisfied by the saturated-model $p$-values of \citet{taylor2014exact}. We derive intuitive and general conditions for independence and show that our proposed constructions yield independent $p$-values. We also discuss connections between the selected-model test and parametric bootstrap sampling.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Many model selection procedures produce a sequence of increasingly complex models, leaving the data analyst to choose among them. 
%While methods like cross-validation aim at minimizing out-of-sample prediction error, such methods tend to select a large fraction of noise variables. 
Given such a path, we consider the problem of choosing the simplest model in the path that is not falsified by the available data. Examples of path algorithms include forward stepwise linear regression, least angle regression (LAR) and  the ever-active path in $\ell_1$-regularized models. \citet{taylor2014exact} study methods for generating exact $p$-values at each step of these path algorithms, and their methods provide a starting point for our proposals. In addition, \citet{choi2014selecting} discuss a selective approach to determining the correct rank of the covariance matrix in principle components analysis.

To select a model, we compute a set of sequential $p$-values at each step of the model path, and then feed them into a stopping rule that is guaranteed to control the FDR, familywise error rate (FWER), or a similar quantity. Recently \citet{gsell2013sequential} and \citet{li2015accumulation} proposed sequential stopping rules of this kind. Both sets of rules require that once we have reached the first correct model in our path, the $p$-values in subsequent steps are uniform (or conservative),
and independent. In this paper we develop a theoretical framework for constructing sequential $p$-values that are both uniform and independent under the correct model, and show how to construct them in forward stepwise regression, LAR and the ever-active path in $\ell_1$-regularized models.

While our approach and analysis are quite general, we begin by introducing a specific example: the selective max-$t$ test for forward stepwise regression, a selective sequential version of the max-$t$ test of \citet{buja2014}, generalized using the theory in \citet{fithian2014optimal}.

\subsection{The max-$t$ Test in Forward Stepwise Regression}

Forward stepwise regression is a greedy algorithm for building a sequence of nested linear regression models. At each iteration, it augments the current model by including the variable that will minimize the residual sum of squares of the next fitted model. Equivalently, it selects the variable with the largest $t$-statistic, controlling for the variables in the current model.

Let $E \sub \{1, \ldots, d\}$ denote the current {\em active set}, the set of predictor variables already selected, and let $\RSS(E)$ denote the residual sum of squares for the corresponding regression model. For $j\notin E$ let $t_{j,E}(Y)$ denote the multivariate $t$-statistic of variable $j$, adjusting for the active variables. Using the standard result that
\begin{equation}
t_{j,E}^2 = (n-|E|-1) \frac{\RSS(E) - \RSS(E \cup \{j\})}{\RSS(E \cup \{j\})},
\end{equation}
we see that the next variable selected is 
\begin{equation}
j^* = \argmax_{j\notin E} |t_{j,E}| = \argmin_{j\notin E} \RSS(E \cup \{j\}),
\end{equation}
with corresponding $t$-statistic $t_E^* = t_{j^*,E}$. The selective max-$t$ test rejects for large values of $t_E^*$, compared to an appropriate conditional distribution that accounts for the adaptive selection of the model path.

Table~\ref{tab:diab} illustrates the max-$t$ test and two others applied to the diabetes data from \cite{LARS}, consisting of observations of $n=442$ patients. The response of interest is a quantitative measure of disease progression one year after baseline, and there are ten measured predictors --- age, sex, body-mass index, average blood pressure, and six blood serum measurements --- plus quadratic terms, giving a total of $d=64$ features. We apply forward stepwise regression to generate the model path (beginning with an intercept term), and then use each of the three methods to generate $p$-values at each step along the path. Finally, we use the ForwardStop rule \citep{gsell2013sequential} at FDR level $\alpha=0.1$ to select a model based on the sequence of $p$-values. The red entry in each column indicates the last variable selected by ForwardStop.

\begin{table}[ht]
\centering
\input{tables/diabetes_edit.tex}
\caption[tab:diab]{\em Forward stepwise regression for the diabetes data: naive $p$-values, $p$-values from the saturated model, and our max-$t$ $p$-values. The red annotation indicates the model selected by ForwardStop with $\alpha=0.1$. For the max-$t$ $p$-values, this stopping rule gives exact $\text{FDR}_{\text{model}}$ control at the 10\% level.}
\label{tab:diab}
\end{table}

The nominal $p$-values in Table~\ref{tab:diab} are computed by comparing $t_E^*$ to the $t_{n-|E|-1}$-distribution, which would be the correct distribution if the sequence of models were selected ahead of time, before observing the data. Because the model sequence is in fact selected adaptively to maximize the value of $t_E^*$, this method is highly anti-conservative.

The max-$t$ test $p$-values are computed using the same test statistic $|t_E^*|$, but compared with a more appropriate null distribution. We can simulate the distribution of $|t_E^*|$ under the null model, i.e. the linear regression model specified by active set $E$:
\begin{equation}
  M(E):\; Y \sim N(X_E\beta, \sigma^2 I), \quad \beta\in \R^{|E|}, \sigma^2>0,
\end{equation}
where $X_E$ is the matrix with columns $(X_j)_{j\in E}$. Because $U_E=\left(X_E'Y,\, \|Y\|^2\right)$ is the complete sufficient statistic for $M(E)$, we can sample from the conditional null distribution of $t_E^*$ given $U_E$ and the current active set $E$, using the {\em selected-model test} framework of \citet{fithian2014optimal}. In step one, because $E=\{1_n\}$ is fixed and $t_E^*$ is independent of $U_E$, the conditional test reduces to the max-$t$ test proposed in \citet{buja2014}. In later steps, $t_E^*$ and $U_E$ are not conditionally independent of $E$, but we can determine the null conditional distribution through Monte Carlo sampling.

The saturated-model $p$-values also account for selection, but they rely on somewhat different assumptions, condition on more information, and test a slightly less restrictive null hypothesis. We discuss these distinctions in detail in Section~\ref{sec:whichnull}.

For most of the early steps in Table~\ref{tab:diab}, the (invalid) nominal test gives the smallest $p$-values, followed by the max-$t$ test, then the saturated-model test. Both the max-$t$ and saturated-model $p$-values are exactly uniform under the null, but the max-$t$ $p$-values appear to be more powerful under the alternative. As we will discuss in Section~\ref{sec:selective-reg}, selected-model tests such as the max-$t$ can be much more powerful than saturated-model tests in early steps of the path when multiple strong variables are compete with each other to enter the model first. Using the max-$t$ $p$-values, ForwardStop selects a model of size 8, compared to size 3 when using the saturated-model $p$-values.

Because the max-$t$ null $p$-values are independent, while the saturated-model $p$-values are not, ForwardStop is guaranteed to control FDR when using the former but not the latter. In Section~\ref{sec:pValsIndep} we discuss intuitive sufficient conditions for independence and show that the max-$t$ and many other selected-model tests satisfy these conditions.

\begin{comment}
The latter two sets of $p$-values test somewhat different hypotheses:  the $p$-values from the saturated model test whether the partial regression of the predictor just added is zero after regressing out the variables already in the model, what we call the ``incremental null''
In contrast, the max-$t$ $p$-values test whether the current model is adequate; that is, whether all predictors not in the model  have coefficients equal to zero, the ``complete null''. (The issue of what hypothesis to test is discussed in section \ref{sec:whichnull}.)
\end{comment}

\subsection{Outline}

For the remainder of the article we will discuss the problem of selective sequential model selection in some generality, returning periodically to the selective max-$t$ test in forward stepwise regression and its lasso regression counterpart, the next-$\lambda$ test, as intuitive and practically important running examples.

In comparison to their saturated-model counterparts, we will see that the selected-model tests proposed here have three main advantages: power, independence (in some cases), and generalizability beyond linear regression. These advantages do not come entirely for free, as we will see: the selected-model methods test a more restrictive null hypothesis than the saturated-model methods.

Section~\ref{sec:genericSetting} introduces the general problem setting, gives important definitions, and reviews relevant literature on testing ordered hypotheses using a sequence of $p$-values. Many of these methods require the $p$-values to be uniform and independent under the null, and we derive general conditions to ensure this property in Section~\ref{sec:pValsIndep}. In Section~\ref{sec:selective-reg} we contrast selected-model and saturated-model tests for linear regression setting, and illustrate why selected-model tests are often much more powerful in early steps of the path. In Section~\ref{sec:computation} we discuss strategies for sampling from the null distribution for our tests, and prove that in forward stepwise regression and $\ell_1$-regularized likelihood methods, the number of constraints in the sampling problem is never more than twice the number of variables. We provide a simulation study in Section~\ref{sec:sparseReg} and briefly discuss applications of our framework to principal components analysis in Section~\ref{sec:pca}. The paper ends with a Discussion.

\section{Selective Sequential Model Selection}
\label{sec:genericSetting}

In the general setting, we observe data $Y \in \cY$ with unknown sampling distribution $F$, and then apply some algorithm to generate an adaptive sequence of $d$ nested statistical models contained in some {\em upper model} $M_{\infty}$.
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By {\em model}, we mean a family of candidate probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration. We will use the notation $M_{0:d}$ to denote the sequence $M_0, \ldots, M_d$, which we call the {\em model path}. We assume throughout that for each candidate model $M$ there is a minimal sufficient statistic $U(Y; \,M)$, and write
\[
U_k(Y) = U(Y; \,M_k(Y)).
\]

Given data $Y$ and model path $M_{0:d}$, we set ourselves the formal goal of selecting the simplest correct model: the smallest $k$ for which $F\in M_k(Y)$. Of course, in most real data problems all of the $M_k(Y)$, as well as all other models under consideration, are incorrect. In more informal terms, then, we seek the simplest model in the path that is not refuted by the available data.

Define the {\em completion index} $k_0(Y) = \min\{k:\; F \in M_k(Y)\}$, the index of the first correct model. By construction, $F\in M_k \iff k \geq k_0$. A {\em stopping rule} is any estimator $\hk(Y)$ of $k_0$, with $M_k$ ``rejected'' if $k < \hk$, and ``accepted'' otherwise. Because $\hk$ is the number of models we do reject, and $k_0$ is the number we should reject, the number of type I errors is $V=(\hk-k_0)_+$, while the number of type II errors is $(k_0-\hk)_+$. Depending on the scientific context, we might want to control the familywise error rate (FWER) $\P(V>0)$, the false discovery rate (FDR) $\E[V/\hk; \hk>0]$ \citep{benjamini1995controlling}, or another error rate defined by the expectation of some loss function $g(\hk, k_0)$:
\begin{equation}\label{eq:errRate}
\Err_F(\hk(\cdot), g) = \E_F\left[ g\left(\hk(Y), k_0(Y)\right)\right]
\end{equation}

We will restrict our attention to stopping rules like those recently proposed by \citet{gsell2013sequential} and \citet{li2015accumulation}, which operate on a sequence $p_{1:d}$ of $p$-values. At each step $k$, we will construct a $p$-value for testing
\[
H_{k}:\; F\in M_{k-1}(Y)
\]
against the alternative that $F\in M_\infty\setminus M_{k-1}$, accounting for the fact that $M_{k-1}$ is chosen adaptively.

Following \citet{fithian2014optimal}, we say that for a fixed candidate model $M$, the random variable $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ at step $k$ if it is conservative under sampling from any $F\in M$, given that $M$ is selected. That is,
\begin{equation}
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\end{equation}

Once we have constructed selective $p$-values for each fixed $M$, we can use them as building blocks to construct a combined $p$-value for the random null $M_{k-1}(Y)$. Define
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is conservative on the event $\{F \in M_{k-1}(Y)\}$:
\begin{equation}\label{eq:selectiveGuaranteePk}
\P_F\left(p_k \leq \alpha \mid F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\end{equation}

One useful strategy for constructing exact selective tests is to condition on the sufficient statistics of the null model $M_{k-1}$. By sufficiency, the law
\[
\L(Y \;\mid\; U_{k-1}, \; M_{k-1}=M)
\]
is the same for every $F\in M$. Thus, we can construct selective tests by comparing the realized value of any test statistic $T(Y)$ to its known conditional distribution under the null. All that remains is to choose the test statistic and compute its conditional null distribution, which can be challenging. See \citet{fithian2014optimal} for a general treatment.

@@

Many familiar path algorithms including lasso, LARS, and forward stepwise regression, are methods for adaptively selecting a set of predictors in linear regression models where we observe a random response $Y\in \R^n$ as well as a fixed design matrix $X\in \R^{n \times d}$, whose columns correspond to candidate predictors. For each possible active set $E \sub \{1,\ldots,d\}$, there is a corresponding candidate model
\[
M(E):\; Y \sim \cN( X_E\beta, \sigma^2 I_n),
\]
which is a subset of the {\em full model}
\[
M_\infty:\; Y \sim \cN(X\beta, \sigma^2I_n).
\]
If the error variance $\sigma^2$ is known, the complete sufficient statistic for $M(E)$ is $X_E'Y$; otherwise it is $\left(X_E'Y,\, \|Y\|^2\right)$.

For the most part, we can discuss our theory and proposals in a generic parametric setting generalizing the linear regression problem above. Let $M_\infty$ be a model parameterized by $\theta\in \Theta \sub \R^{\cJ}$:
\[
M_\infty = \{F_\theta:\; \theta \in \Theta\}.
\]
For any subset $E\sub \cJ$ define the sparse submodel with active set $E$ as follows:
\[
\Theta(E) = \{\theta:\; \theta_j = 0, \;\;\forall j \notin E\}, 
\quad M(E) = \{F_\theta:\; \theta\in \Theta(E)\}.
\]

We can consider path algorithms that return a sequence of nested active sets
\[
E_0(Y) \sub E_1(Y) \sub \cdots \sub E_d(Y) \sub \cJ,
\]
inducing a model path with $M_k = M(E_k),$ for $k=0,\ldots,d$. We will be especially interested in two generic path algorithms for the sparse parametric setting: forward stepwise paths and ever-active regularization paths.

\subsubsection{Forward Stepwise Paths and Greedy Likelihood Ratio Tests}
Let $\ell(\theta; Y)$ denote the log-likelihood for model
$M_\infty$. The {\em forward stepwise} algorithm proceeds as follows: we begin with some fixed $E_0$ (such as an intercept-only model), then at step $k=1,\ldots,d$, we set
\begin{equation}
j_k = \argmax_j \;\;\sup \left\{\ell(\theta; Y):\; \theta\in\Theta(E_{k-1} \cup \{j\})\right\}, \quad \text{ and } \quad
E_k = E_{k-1} \cup \{j_k\}.
\end{equation}
That is, at each step we select the next variable to maximize the likelihood of the next fitted model.

A natural choice of test statistic is the {\em greedy likelihood ratio statistic}
\begin{equation}\label{eq:greedyLRT}
G_k(Y) = \sup_{\theta\in \Theta(E_k)} {\ell(\theta; Y)} \;\;- \sup_{\theta\in \Theta(E_{k-1})} {\ell(\theta; Y)},
\end{equation}
the generalized likelihood ratio statistic for testing $M(E_{k-1})$ against the ``greedy'' alternative with one more active parameter:
\begin{equation}\label{eq:greedyAlternative}
M(E_{k-1})^{+1} = \bigcup_{j \notin E_{k-1}} M(E_{k-1} \cup \{j\}).
\end{equation}

We define the {\em selective greedy likelihood ratio test} as the test that rejects for large $G_k$, based on the law
\begin{equation}
\L\left(G_k \;\mid\; M_{k-1}, U_{k-1}\right)
\end{equation}

Because the likelihood in linear regression is a monotone decreasing function of the residual sum of squares, the max-$t$ test is equivalent to the greedy likelihood test.

For simplicity, we have made two assumptions: that only one variable is added at each step, and that the set of candidate variables we choose from is the same in each step. It is relatively straightforward to relax either assumption, as we discuss briefly in Section~\ref{sec:discussion}.

\subsubsection{Ever-Active Regularization Paths and next-$\lambda$ Tests}
Another important class of model selection procedures is the sequence of {\em ever-active} sets for a regularized likelihood path. For some ordered set $\Lambda$, let $\lambda\in\Lambda$ parameterize the regularization penalty $P_\lambda(\theta)$. In the case of lasso-penalized regression, we could take $P_\lambda = \lambda\|\beta\|_1$ for $\lambda\in (0,\infty]$.

We then define
\begin{align}
  \hat\theta^{\lambda}(Y) &= 
  \argmin_{\theta\in\Theta} -\ell(\theta; Y) + P_\lambda(\theta) \\
  \tE_\lambda(Y) &= \left\{j:\; \hat\theta_j^\nu \neq 0 
    \text{ for any } \nu\in \Lambda, \nu \geq \lambda \right\}
\end{align}
It may be impossible or inconvenient to compute $\hat\theta^\lambda$ for every $\lambda\in(0,\infty]$. If so, we can instead take $\Lambda$ to be a grid of finitely many positive values.
